---
title: "US Flight Delays Analysis "
author: "Markus Köfler"
date: "2022-09-21"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Introduction

This project is intended to demonstrate the skills acquired from the Google Data Analytics Certificate Course hosted on Coursera. The dataset was retrived from [Kaggle](%22https://www.kaggle.com/datasets/undersc0re/flight-delay-and-causes/metadata?select=Flight_delay.csv%22). Originally, the dataset comes form the [U.S. Department of Transportation's (DOT) Bureau of Transportation Statistics (BTS)](https://www.bts.gov/).

# The Business Task

1) A business consultancy company is sending their consultants to their customers within the US area (domestic flights). The head of the HR team wants to know what are the most reliable airlines, what airports cause the most frequent and longest delays and cause the most frequent and longest delays in order to cut costs and minimize travel times for employees.

2) While working on the analysis, the CEO wants to find out [clients should be served on time ]

3) The consultancy company is located in Chicago (IL), Austin (TX), Denver (CL) , Jersey City (NJ) and Seattle (WA)



















This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Slide with Bullets

-   Bullet 1
-   Bullet 2
-   Bullet 3

## Slide with R Output

```{r cars, echo = TRUE}
summary(cars)
```

## Slide with Plot

```{r pressure}
plot(pressure)
```
